### 2022-05-08 MSC-GUDMUNDSSON-2022.tex
General fleshing out, refinement and fixes.

%% An abstract is required to be in both Icelandic and English for most degrees.
%% It is considered good form to limit the abstract to a single paragraph in each language,
%%   at 300 words.  Refer to your degree's instructions.
%% Note: Icelandic quotation marks cannot be typeset using "` and "'.  You should use \enquote{}
%% this is probably due to interactions with the MultiLingual macros.
%% TODO: turn this into more sensible macros to avoid confusion --foley
\newML{AbstractText}{
The world is moving at increasing speed away from generating electricity using fossil fuels, toward more environmentally friendly options, such as solar power. Solar power production is volatile and highly dependent on weather conditions, primarily cloud cover. Instability and unpredictability are undesirable in power production since production must match demand. In order to ensure power availability when dependent on solar power, highly polluting and expensive fossil power stations, \enquote{peaker plants}, are kept on standby to react to the instability in solar power. The ability to reliably predict solar power production provides more predictability to power grid operators, allowing for greater efficiency in grid operation. 
Valuable data exists in weather forecasts, which can be harnessed to increase predictability in solar power generation. We use novel deep learning methods, attention, and specifically Temporal Fusion Transformers, to forecast the irradiance at a solar power station, which correlates very strongly with power output. We base this on Belgingur ehf.'s high-resolution weather forecast data, the 24-hour forecast irradiance values for the area of the power station.
Our model achieves a very low average Mean Absolute Percentage Error of 13.21\% on the test dataset for one-day forecasts. In addition, the model outputs a confidence level for its predictions providing insight to a grid operator on whether to expect unpredictability in the power output. The model is able to generate more accurate predictions when utilizing forecast data from a wider area around the power station than exclusively data from the power station‘s immediate vicinity.

}  
{
Hraðar og hraðar erum við að færast frá því að framleiða raforku með jarðefnaeldsneyti, í átt að umhverfisvænni mátum til raforkuframleiðslu eins og sólarorku. Sólarorka er óútreiknanleg og mjög háð veðurfari, aðallega skýjahulu. Óstöðugleiki og óútreiknanleiki eru óákjósanlegir þegar kemur að sólarorku, þar sem framleiðsla verður að vera jöfn neyslu. Til að tryggja afhendingu raforku, þegar sólarorka er notuð, reiða framleiðendur sig oft á dýr og mengandi orkuver sem ganga fyrir jarðefnaeldsneyti og standa í viðbragðsstöðu til að bregðast við óvæntum breytingum á sólarorkuframboði. Áreiðanleg geta til að sjá fyrir um sólarorkuframleiðslu gerir stjórnun raforkukerfa einfaldari og útreiknanlegri, sem eykur skilvirkni þeirra.
Verðmæt gögn eru til í formi veðurspáa, sem hægt er að nýta til að gera sólarorkuframleiðslu útreiknanlegri. Við notum nýjar aðferðir á sviði djúpnáms, athygli(e. attention), og þá sérstaklega tímasamrunabylta (e. temporal fusion transformers) til að spá fyrir um sólargeislun við sólarorkuver, sem fylgir náið raforkuframleiðslu orkuversins. Við byggjum þessa vinnu á háupplausnarveðurspám Belgings ehf., 24 klst. geislunarspám fyrir staðsetningu okuversins.
Módelið nær meðal beinni meðalprósentuskekkju (e. mean absolute percentage error) upp á 13.21\% á prófunargagnasettinu fyrir eins dags spár. Að auki skilar módelið öryggisbili fyrir spárnar, sem veitir raforkuframleiðendum innsæi um það hvort þeir geti búist við óútreiknanlegri sólarorkuframleiðslu. Módelið skilar nákvæmari spám þegar það styðst við veðurspágögn af stærra svæði í nánd orkuversins en þegar það nýtir eingöngu gögn fyrir nærumhverfi orkuversins.
} % Icelandic abstract goes here
\setAbstract{\MLAbstractText}

@@

%% Dedication is optional, comment out if it is absent
%% RUM: Not mentioned
\begin{dedications}
  I dedicate this to my spouse/child/pet/power animal.
\end{dedications}

@@

\draftnote{Acknowledgements are optional; comment this chapter out if they are absent
  Note that it is important to acknowledge any funding that helped in the work}

@@

%% The list of abbreviations is an example of a special list
%% Other lists may be added, such as lists of algorithms, symbols, theorems, etc.
%% IN CS PhD, this is sometimes centered.
\coverchapter{List of Abbreviations}%%RUM: Not mentioned
\begin{tabular}{ll}
MSc &Masters of Science\\
PhD &Doctor of Philosophy\\
AI &Artificial Intelligence\\
ANN &Artificial Neural Network\\
NN &Neural Network\\
RNN &Recurrent Neural Network\\
GRU &Gated Recurrent Unit\\
LSTM &Long Short Term Memory\\
NLP &Natural Language Processing\\
TCNN &Temporal Convolutional Neural Network\\
NARXNN &Non-linear Auto Regressive Neural Network\\
TFT &Temporal Fusion Transformer\\
GRN &Gated Residual Network\\
MAPE &Mean Absolute Percentage Error\\
MAE &Mean Absolute Error\\
CUDA &Compute Unified Device Architecture\\
AMD &Advanced Micro Devices inc.\\
CPU &Central Processing Unit\\
RAM &Random Access Memory\\

>>>>

%% An abstract is required to be in both Icelandic and English for most degrees.
%% It is considered good form to limit the abstract to a single paragraph in each language,
%%   at 300 words.  Refer to your degree's instructions.
%% Note: Icelandic quotation marks cannot be typeset using "` and "'.  You should use \enquote{}
%% this is probably due to interactions with the MultiLingual macros.
%% TODO: turn this into more sensible macros to avoid confusion --foley
\newML{AbstractText}{
The world is moving at increasing speed away from generating electricity using fossil fuels, toward more environmentally friendly options, such as solar power. Solar power production is volatile and highly dependent on weather conditions, primarily cloud cover. Instability and unpredictability are undesirable in power production since production must match demand. In order to ensure power availability when dependent on solar power, highly polluting and expensive fossil power stations, \enquote{peaker plants}, are kept on standby to react to the instability in solar power. The ability to reliably predict solar power production provides more predictability to power grid operators, allowing for greater efficiency in grid operation. 
Valuable data exists in weather forecasts, which can be harnessed to increase predictability in solar power generation. We use novel deep learning methods, attention, and specifically temporal fusion transformers, to forecast the irradiance at a solar power station, which correlates very strongly with power output. We base this on Belgingur ehf.'s high-resolution weather forecast data, the 24-hour forecast irradiance values for the area of the power station.
Our model achieves a very low average mean absolute percentage error of 13.21\% on the test dataset for one-day forecasts. In addition, the model outputs a confidence level for its predictions providing insight to a grid operator on whether to expect unpredictability in the power output. The model is able to generate more accurate predictions when utilizing forecast data from a wider area around the power station than exclusively data from the power station‘s immediate vicinity.

}  
{
Hraðar og hraðar erum við að færast frá því að framleiða raforku með jarðefnaeldsneyti, í átt að umhverfisvænni mátum til raforkuframleiðslu eins og sólarorku. Sólarorka er óútreiknanleg og mjög háð veðurfari, aðallega skýjahulu. Óstöðugleiki og óútreiknanleiki eru óákjósanlegir þegar kemur að sólarorku, þar sem framleiðsla verður að vera jöfn neyslu. Til að tryggja afhendingu raforku, þegar sólarorka er notuð, reiða framleiðendur sig oft á dýr og mengandi orkuver sem ganga fyrir jarðefnaeldsneyti og standa í viðbragðsstöðu til að bregðast við óvæntum breytingum á sólarorkuframboði. Áreiðanleg geta til að sjá fyrir um sólarorkuframleiðslu gerir stjórnun raforkukerfa einfaldari og útreiknanlegri, sem eykur skilvirkni þeirra.
Verðmæt gögn eru til í formi veðurspáa, sem hægt er að nýta til að gera sólarorkuframleiðslu útreiknanlegri. Við notum nýjar aðferðir á sviði djúpnáms, athygli(e. attention), og þá sérstaklega tímasamrunabylta (e. temporal fusion transformers) til að spá fyrir um sólargeislun við sólarorkuver, sem fylgir náið raforkuframleiðslu orkuversins. Við byggjum þessa vinnu á háupplausnarveðurspám Belgings ehf., 24 klst. geislunarspám fyrir staðsetningu okuversins.
Módelið nær meðal meðaltölugildisprósentuskekkju (e. mean absolute percentage error) upp á 13,21\% á prófunargagnasettinu fyrir eins dags spár. Að auki skilar módelið öryggisbili fyrir spárnar, sem veitir raforkuframleiðendum innsæi um það hvort þeir geti búist við óútreiknanlegri sólarorkuframleiðslu. Módelið skilar nákvæmari spám þegar það styðst við veðurspágögn af stærra svæði í nánd orkuversins en þegar það nýtir eingöngu gögn fyrir nærumhverfi orkuversins.
} % Icelandic abstract goes here
\setAbstract{\MLAbstractText}

@@

%% Dedication is optional, comment out if it is absent
%% RUM: Not mentioned
\begin{dedications}
  I dedicate this to my dog Birta. Rest in peace.
\end{dedications}

@@

I would like to thank my supervisor, Yngvi, for invaluable advice throughout this work and Ólafur and Jorge at Belgingur for their help with understanding the weather and solar power aspect of the problem. I would also like to thank my friends, family and fellow thesis writers for being around and keeping me company while I was working on this.\\

@@

%% The list of abbreviations is an example of a special list
%% Other lists may be added, such as lists of algorithms, symbols, theorems, etc.
%% IN CS PhD, this is sometimes centered.
\coverchapter{List of Abbreviations}%%RUM: Not mentioned
\begin{tabular}{ll}
MSc &Master of Science\\
PhD &Doctor of Philosophy\\
AI &Artificial Intelligence\\
ANN &Artificial Neural Network\\
NN &Neural Network\\
RNN &Recurrent Neural Network\\
GRU &Gated Recurrent Unit\\
LSTM &Long Short Term Memory\\
NLP &Natural Language Processing\\
TCNN &Temporal Convolutional Neural Network\\
NARXNN &Non-linear Auto Regressive Neural Network\\
TFT &Temporal Fusion Transformer\\
GRN &Gated Residual Network\\
MAPE &Mean Absolute Percentage Error\\
MAE &Mean Absolute Error\\
CUDA &Compute Unified Device Architecture\\
AMD &Advanced Micro Devices inc.\\
CPU &Central Processing Unit\\
RAM &Random Access Memory\\

### 2022-05-08 background.tex
General fleshing out, refinement and fixes.

\chapter{Background\label{cha:background}}
In this chapter we discuss the work our work builds on to give our work context. We cover the history of neural networks in machine learning, previous architectures that have been used to solve similar problems and the components of the solution we arrived at.

\section{Neural Networks}
Artificial Neural Networks (ANNs), generally known as just Neural Networks (NNs) were first theorized in the 1940s \cite{mcculloch_logical_1943}. They are supposed to emulate the function of brains. They have \enquote{neurons} which, when they receive a signal, apply a function to the signal which they relay to the next neuron. Neural networks are built in layers. A layer receives a vector of input values, each neuron in the layer takes a weighted sum of the inputs and applies its activation function to the result. These values are then passed on to the next layer. Training, a neural network is the process of optimizing the weights in the layers of the network such that a specified input yields the desired output \cite{goodfellow_deep_2016}.\\
Frank Rosenblatt Perceptron in 1950s. A single layer of neurons with the ability to classify images a few hundred pixels in size. Train neurons based on a dataset. His training algorithm was very limited and so inefficient that it did not allow for neural networks with more than one layer \cite{rosenblatt_perceptron_1958}. First theorized ability to recognize people and translate between languages.\\
In 1974 Paul Werbos proposed backpropagation to optimize neural networks \cite{Werbos:74}, which became popular in 1986  with the work of Rumelhart et al..  Backpropagation is an efficient training algorithm opening up the possibility of training neural networks deeper than one layer using gradient methods\cite{noauthor_backpropagation_2022}. Backpropagation enabled the creation of deeper neural networks, notably used to read handwritten digits with good accuracy \cite{lecun_backpropagation_1989}.\\
Although backpropagation enabled neural networks with multiple layers, it did not enable significant depth. The most commonly used activation functions of the time, sigmoid and tanh, have derivatives close to zero, the gradient in the network values flatten at an exponential rate. This slows down the gradient descent used to optimize the networks, also at an exponential rate. In the 2010s the ReLu activation function \cite{nair_rectified_2010} coupled with improved optimization algorithms like ADAM and massive advances in computer hardware opened the door for Deep Learning, neural networks with many more layers. Today, deep neural networks have been used to great effect in applications like Google DeepMind's AlphaGO \cite{silver_mastering_2016}, the first AI to beat the world champion in the game of GO, and OpenAI's GPT3 model generates text human-like enough to pass the Turing test, to be confused with text written by a human \cite{vandeput_brief_nodate}.

\section{LSTMs and GRUs\label{cha:RNN}}
Recurrent Neural Networks (RNNs), especially Gated Recurrent Units (GRUs) and Long-Short Term Memory (LSTM) cells \cite{Goodfellow-et-al-2016} along with more advanced derivatives have been shown to be able to achieve good results when trained on previous weather and power generation data \cite{lin_temporal_2020, lee_forecasting_2018, jaidee_very_2019, su_machine_2019}.
RNNs are a class of deep neural networks where nodes are connected in a temporal sequence, allowing the network to learn time-based data efficiently\cite{lim_time_2021}. LSTMs and GRUs are types of RNNs in which, the nodes in the network have additional stored states controlled by the network via mechanisms that use time delays or feedback loops \cite{Goodfellow-et-al-2016, noauthor_recurrent_2021}.\\

@@

\section{Transformers\label{cha:transformer}}
%Hard to train
%sequential
% -> parallelization hard
%encoder/decoder instead. Attention. 
% Explain enc/dec
% Autoencoders generate same output as the input, so no labeled data is needed -> completely unsupervisd earning.
% Locality a problem with CNNs and RNNs. Can easily find local context but keeping on to context far apart in input is difficult. Enter transformers.
 %Encoder reads blocks of data and uses that and previously generated data as inputs into decoder which generatesdata one point at a time.
 %Attention. How you relate things to each other. i.e. input to output
 %self attention. How do features in the input data relate to each other
 \textcolor{red}{TODO: Make this coherent}

Transformers have been used for a while in Natural Language Processing (NLP) and have more recently been used very effectively in forecasting \cite{vaswani_attention_2017}. They can more effectively learn and preserve local context located far apart in the input data. As an example from NLP, the training data can contain multiple cases where time is being told, far apart. A transformer model would then more easily identify that the word “o’clock” is generally preceded by a number, whereas RNNs would be more likely to struggle with such distributed information. This can be applied to our work as similar weather patterns would reoccur days apart and cause a similar effect on measured irradiance at the power station \cite{lim_time_2021}.

\afterpage{%
\begin{figure}[ht!]
    \centering
    \includegraphics[scale=0.95]{imgs/transformer.png}
    \caption{The transformer architecture.\cite{vaswani_attention_2017}
    \label{fig:transformer}}
\end{figure}
\clearpage
}

\subsection{Encoder/decoder}
Transformers have an encoder/decoder structure. An encoder takes an input sequence and generates an abstract hidden state which the decoder then uses to generate an output sequence. A major advantage of this methodology is that the inputs and outputs do not have to have the same length, whereas the primary limit of the architecture is that all information in the input sequence needs to be represented by a single vector for the decoder to decode. The encoder distils the input down to a much smaller representation, a bottleneck, only containing desired information which the decoder in turn extracts. This is on its own used for tasks like de-noising or colourizing images.

\begin{figure}[ht!]
    \centering
    \includegraphics[scale=0.5]{imgs/encoder_decoder.jpeg}
    \caption{The encoder/decoder structure \cite{nechu_what_2020}
    \label{fig:encoder_decoder}}
\end{figure}

\subsection{Positional encoding}
Unlike RNNs, the transformer architecture has no recurrence and does inherently track the order of the input data, so positional information has to be injected into the network input for the model to be able to use the positional context of the data.\cite{vaswani_attention_2017}

@@

\section{Previous work}
Lin et al. \cite{lin_temporal_2020} showed that while the more traditional GRUs and LSTMs show good performance on the task, Temporal Convolutional Neural Networks (TCNN) can show even better performance. \textit{"TCNN is a novel convolutional architecture designed for sequential modelling, which combines causal and dilated convolutions
and residual connections"} \cite{lin_temporal_2020}. \\
Jaidee et al. \cite{jaidee_very_2019} showed that LSTMs, GRUs and derivative methods show relatively similar performance for very short-term predictions on the time scale of a few hours. This timescale is where it is critical to accurately predict cloud movements. 
These methods have been taken as far as building genetic algorithms to generate the optimal neural network for the purpose \cite{jaidee_very_2019}. These genetic algorithms operate on the same principles as animal breeding. They generate multiple candidate neural networks and use the best networks as a base to generate new networks for the next generation. This is done for multiple generations and in the end, the best neural network generated is used \cite{jaidee_very_2019}.\\
Su et al. \cite{su_machine_2019} did extensive experimentation on more conventional neural networks and newer, less known networks including  Non-linear Auto Regressive Neural Network (NARXNN), as well as various statistical approaches. When predicting power output, NARXNN takes in the current power output values and considers them as well as the past values of the power output of the system. These experiments showed that NARXNN had the best performance of these, by a rather large margin, and a hybrid method, of the better performing methods, showed even better results \cite{anderson_using_2018}. The statistical methods generally performed significantly worse in predicting the power output. \\
Satellite imagery has been used for predicting solar power output as well. A neural network was trained to predict solar power production, primarily from cloud cover information it learned from these satellite images \cite{jang_solar_2016}.\\

>>>>

\chapter{Background\label{cha:background}}
In this chapter, we discuss the work our work builds on to give our work context. We cover the history of neural networks in machine learning, previous architectures that have been used to solve similar problems and the components of the solution we arrived at.

\section{Neural Networks}
Artificial Neural Networks (ANNs), generally known as just Neural Networks (NNs) were first theorized in the 1940s \cite{mcculloch_logical_1943}. They are supposed to emulate the function of biological brains. They are composed of \enquote{neurons} which, when they receive a signal, apply a function to the signal which they then relay to the next neuron. Neural networks are built in layers. A layer receives a tensor of input values, each neuron in the layer takes a weighted sum of the inputs and applies its activation function to the result. These values are then passed on to the next layer. Training a neural network is the process of optimizing the weights in the layers of the network such that a specified input yields the desired output via iteration to minimize the output of a loss function \cite{goodfellow_deep_2016}.\\
Frank Rosenblatt created the perceptron in the 1950s. A perceptron is a single layer of neurons with the ability to classify images a few hundred pixels in size. Rosenblatt's breakthrough idea was to train a neural network based on reinforcement from a dataset, strengthening successful connections and weakening unsuccessful connections. His training algorithm was very limited and so inefficient that it did not allow for neural networks with more than one layer \cite{rosenblatt_perceptron_1958}. Based on his work, Rosenblatt first theorized the ability of neural networks to recognize people and translate between languages \cite{vandeput_brief_nodate}.\\
In 1974 Paul Werbos proposed backpropagation to optimize neural networks \cite{Werbos:74}, which became popular in 1986  with the work of Rumelhart et al. \cite{rumelhart_learning_1986}.  Backpropagation is an efficient training algorithm opening up the possibility of training neural networks deeper than one layer using gradient descent. \enquote{The backpropagation algorithm works by computing the gradient of the loss function with respect to each weight by the chain rule, computing the gradient one layer at a time, iterating backward from the last layer to avoid redundant calculations of intermediate terms in the chain rule} \cite{noauthor_backpropagation_2022}. Backpropagation enabled the creation of deeper neural networks, notably used to read handwritten digits with good accuracy \cite{lecun_backpropagation_1989}.\\
Although backpropagation enabled neural networks with multiple layers, it did not enable significant depth. The most commonly used activation functions of the time, sigmoid and tanh, have derivatives close to zero, so stacking them flattens the gradient in the network values at an exponential rate. This slows down the gradient descent used to optimize the networks, also at an exponential rate. In the 2010s the ReLu activation function \cite{nair_rectified_2010} coupled with improved optimization algorithms like ADAM and massive advances in computer hardware opened the door for Deep Learning, neural networks with many more layers. Today, deep neural networks have been used to great effect in applications like Google DeepMind's AlphaGO \cite{silver_mastering_2016}, the first AI to beat the world champion in the game of go, and OpenAI's GPT3 model generates text human-like enough to pass the Turing test, to be confused with text written by a human \cite{vandeput_brief_nodate}.

\section{Recurrent Neural Networks\label{cha:RNN}}
Recurrent Neural Networks (RNNs), especially Gated Recurrent Units (GRUs) and Long Short-Term Memory (LSTM) cells \cite{Goodfellow-et-al-2016, hochreiter_long_1997} along with more advanced derivatives have been shown to be able to achieve good results when trained on previous weather and power generation data \cite{lin_temporal_2020, lee_forecasting_2018, jaidee_very_2019, su_machine_2019}.
RNNs are a class of deep neural networks where nodes are connected in a temporal sequence, allowing the network to learn time-based data efficiently \cite{lim_time_2021}. LSTMs and GRUs are types of RNNs in which, the nodes in the network have additional stored states controlled by the network via mechanisms that use time delays or feedback loops \cite{Goodfellow-et-al-2016, noauthor_recurrent_2021}. Figure~\ref{fig:RNNs} shows the structure of the nodes of the three different Recurrent Neural Network types discussed above.\\

\begin{figure}[ht!]
    \centering
    \subfloat[Plain Recurrent Neural Network (RNN)]{\includegraphics[scale=0.35]{imgs/RNN.png}\label{fig:RNN}}\qquad
    \subfloat[Gated Recurrent Unit (GRU)]{\includegraphics[scale=0.35]{imgs/GRU.png}\label{fig:GRU}}\qquad
    \subfloat[Long Short-Term Memory (LSTM)]{\includegraphics[scale=0.35]{imgs/LSTM.png}\label{fig:LSTM}}\qquad
    \caption{The cells of different Recurrent Neural Network types. $x_t$ represents inputs, $h_t$ and $C_t$ represent hidden layer vectors passed between cells. $\sigma$ and $tanh$ are activation functions. \cite{lopez_rnn_nodate}.
    \label{fig:RNNs}}
\end{figure}

@@

\section{Transformers\label{cha:transformer}}
%Hard to train
%sequential
% -> parallelization hard
%encoder/decoder instead. Attention. 
% Explain enc/dec
% Autoencoders generate same output as the input, so no labeled data is needed -> completely unsupervisd earning.
% Locality a problem with CNNs and RNNs. Can easily find local context but keeping on to context far apart in input is difficult. Enter transformers.
 %Encoder reads blocks of data and uses that and previously generated data as inputs into decoder which generatesdata one point at a time.
 %Attention. How you relate things to each other. i.e. input to output
 %self attention. How do features in the input data relate to each other

Transformers have been used for a while in Natural Language Processing (NLP) and have more recently been used very effectively in forecasting \cite{vaswani_attention_2017}. They can more effectively learn and preserve local context located far apart in the input data. As an example from NLP, the training data can contain multiple cases where time is being told, far apart. A transformer model would then more easily identify that the word \enquote{o’clock} is generally preceded by a number, whereas RNNs would be more likely to struggle with such distributed information. This can be applied to our work as similar weather patterns would reoccur days apart and cause a similar effect on measured irradiance at the power station \cite{lim_time_2021}. Figure~\ref{fig:transformer} shows an overview of the transformer architecture.

\afterpage{%
\begin{figure}
    \centering
    \includegraphics[scale=0.80]{imgs/transformer.png}
    \caption{The transformer architecture.\cite{vaswani_attention_2017}
    \label{fig:transformer}}
\end{figure}
\clearpage
}

\subsection{Encoder/decoder}
Transformers have an encoder/decoder structure, depicted in Figure~\ref{fig:encoder_decoder}. An encoder takes an input sequence and generates an abstract hidden state which the decoder then uses to generate an output sequence. A major advantage of this methodology is that the inputs and outputs do not have to have the same length, whereas the primary limit of the architecture is that all information in the input sequence needs to be represented by a single vector for the decoder to decode. The encoder distils the input down to a much smaller representation, a bottleneck, only containing desired information which the decoder in turn extracts. This is on its own used for tasks like de-noising or colourizing images.

\begin{figure}
    \centering
    \includegraphics[scale=0.5]{imgs/encoder_decoder.jpeg}
    \caption{The encoder/decoder structure \cite{nechu_what_2020}
    \label{fig:encoder_decoder}}
\end{figure}

\subsection{Positional encoding}
Unlike RNNs, the transformer architecture has no recurrence, or cell based memory, and does not inherently track the order of the input data, so positional information has to be injected into the network input for the model to be able to use the positional context of the data.\cite{vaswani_attention_2017}

@@

\section{Previous work}
Lin et al. \cite{lin_temporal_2020} showed that while the more traditional GRUs and LSTMs show good performance on the task, Temporal Convolutional Neural Networks (TCNN) can show even better performance. \textit{"TCNN is a novel convolutional architecture designed for sequential modelling, which combines causal and dilated convolutions
and residual connections"} \cite{lin_temporal_2020}. \\
Jaidee et al. \cite{jaidee_very_2019} showed that LSTMs, GRUs and derivative methods show relatively similar performance for very short-term predictions on the time scale of a few hours. This timescale is where it is critical to accurately predict cloud movements. 
These methods have been taken as far as building genetic algorithms to generate the optimal neural network for the purpose \cite{jaidee_very_2019}. These genetic algorithms operate on the same principles as animal breeding. They generate multiple candidate neural networks and use the best networks as a base to generate new networks for the next generation. This is done for multiple generations and in the end, the best neural network generated is used \cite{jaidee_very_2019}.\\
Su et al. \cite{su_machine_2019} did extensive experimentation on more conventional neural networks and newer, less known networks including  Non-linear Auto Regressive Neural Networks (NARXNN), as well as various statistical approaches. When predicting power output, NARXNN takes in the current power output values and considers them, in real time, as well as the past values of the power output of the system. These experiments showed that NARXNN had the best performance of these, by a rather large margin, and a hybrid method, of the better performing methods, showed even better results \cite{anderson_using_2018}. The statistical methods generally performed significantly worse in predicting the power output. \\
Satellite imagery has been used for predicting solar power output as well. A neural network was trained to predict solar power production, primarily from cloud cover information it learned from these satellite images \cite{jang_solar_2016}.\\
\textcolor{red}{AlexNet}

### 2022-05-08 discussion.tex
Implemented feedback from instructor
General fleshing out, refinement and fixes.

\section{Conclusion\label{sec:conclusions}}
The results clearly show that relying on forecast irradiance data from meteorological models is a very effective, but not perfect, method of accurately predicting irradiance in a single area. A transformer model, the Temporal Fusion Transformer, managed to very effectively translate patterns in meteorological irradiance forecasts to accurate local irradiance predictions utilizing spatially and temporally adjacent data.

We used a dataset of a fairly limited size, only 425 days. A few years of data for training the model would likely yield somewhat better results. The data can be processed more before being input into the network to account for known variables like daily and yearly swings in irradiance, which might improve performance. There exists satellite irradiance measurement data that might be used for improving accuracy. It would be interesting to see if a neural network could extract information from the seven pyranometers scattered across the power station, rather than only dropping the highest and lowest values and taking an average.

Sadly, there is not a standard benchmark for solar irradiance forecast models, and the data used by others who have done similar work is not available, so we can't make a proper comparison between methodologies and models. 
%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "DEGREE-NAME-YEAR"
%%% End:

>>>>

\section{Limitations and Future Work}
We used a dataset of a fairly limited size, only 425 days. A few years of data for training the model would likely yield somewhat better results. The data can be processed more before being input into the network to account for known variables like daily and yearly swings in irradiance, which might improve performance. There exists satellite irradiance measurement data that might be used for improving accuracy. It would be interesting to see if a neural network could extract information from the seven pyranometers scattered across the power station, rather than only dropping the highest and lowest values and taking an average.

The model also has huge potential in utilizing more varied data in addition to a larger time span, other exogenous data that might further help relate the input forecast to the actual irradiance. Experimenting with wind data, both historical and forecast data, may well unlock further steps in performance.

Sadly, there is not a standard benchmark for solar irradiance forecast models, and the data used by others who have done similar work is not available, so we can't make a proper comparison between methodologies and models.

\section{Conclusion\label{sec:conclusions}}
The results clearly show that relying on forecast irradiance data from meteorological models is a very effective, but not perfect, method of accurately predicting irradiance in a single area. A transformer model, the Temporal Fusion Transformer, managed to very effectively translate patterns in meteorological irradiance forecasts to accurate local irradiance predictions utilizing spatially and temporally adjacent data.\\
%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "DEGREE-NAME-YEAR"
%%% End:

### 2022-05-08 introduction.tex
General fleshing out, refinement and fixes.

Solar power production is highly dependent on weather conditions \cite{lin_temporal_2020, lee_forecasting_2018, jaidee_very_2019, su_machine_2019, jang_solar_2016}. A single cloud can adversely impact the production of a solar farm. Such unreliability in a power production system is undesirable since production must match demand \cite{lee_forecasting_2018}. When solar power production falls, the power demand it was serving must be fulfilled by other means. Fossil fueled power stations are generally kept on standby, waiting to fulfil this demand \cite{lee_forecasting_2018}. These power stations often take hours to start, generally pollute more than other power stations and are expensive to run \cite{lee_forecasting_2018}.
Reliable and accurate forecasts of the fluctuations in solar power production can give power grid operators notice and insight into future power production \cite{lee_forecasting_2018}. This notice enables planning for the dips in power production and making arrangements for the future. This provides the tools to reduce the cost and pollution of power generation as well as increase the reliability of power delivery to consumers \cite{lee_forecasting_2018}.
Recent advancements in deep learning have unlocked new tools that have been shown to be effective in predicting the future behaviour of a system given enough relevant data.\\

@@

\section{Goals}
Previous methods utilized in the prediction of solar power production all have in common that they must effectively predict the weather so they can predict power output.

Here we explore using high-resolution weather forecast data from meteorological models, offloading a majority of this difficult task to the field of meteorology. Instead, we focus on correlating these forecasts with measurements and detecting and reducing temporal and spatial error in the general forecasts using newer methods like Temporal Fusion Transformers.
What is the quality of the predictions this methodology produces? Can we create a model which is useful to power producers in operating their systems more efficiently?

The thesis is structured as follows. First we talked about the motivations for the work. Next we discuss the background the work stands on. Next we cover the methods utilized, the data and tools used to achieve these goals. The proceeding chapter contains the results of the work, and finally we discuss the implications, strengths and flaws of the work, future work and end with conclusions.

>>>>

\vspace{1.5cm}

Solar power production is highly dependent on weather conditions \cite{lin_temporal_2020, lee_forecasting_2018, jaidee_very_2019, su_machine_2019, jang_solar_2016}. A single cloud can adversely impact the production of a solar farm. Such unreliability in a power production system is undesirable since production must match demand \cite{lee_forecasting_2018}. When solar power production falls, the power demand it was serving must be fulfilled by other means. Fossil fueled power stations are generally kept on standby, waiting to fulfil this demand \cite{lee_forecasting_2018}. These power stations often take hours to start, generally pollute more than other power stations and are expensive to run \cite{lee_forecasting_2018}.
Reliable and accurate forecasts of the fluctuations in solar power production can give power grid operators notice and insight into future power production \cite{lee_forecasting_2018}. This notice enables planning for the dips in power production and making arrangements for the future. This provides the tools to reduce the cost and pollution of power generation as well as increase the reliability of power delivery to consumers \cite{lee_forecasting_2018}.
Recent advancements in deep learning have unlocked new tools for forecasting, such as attention mechanisms, that have been shown to be effective in predicting the future behaviour of a system, given enough relevant data.\\

@@

 \newpage
\section{Goals}
Previous methods utilized in the prediction of solar power production all have in common that they must effectively predict the weather so they can predict power output.

Here we explore using high-resolution weather forecast data from meteorological models, offloading a majority of this difficult task to the field of meteorology. Instead of generating forecasts, we focus on correlating existing forecasts with measurements and detecting and reducing temporal and spatial error in the general forecasts using newer methods like Temporal Fusion Transformers.
What is the quality of the predictions this methodology produces? Can we create a model which is useful to power producers in operating their systems more efficiently?

This thesis is structured as follows. First, we talked about the motivations for the work. Next, we discuss the background the work stands on. After that, we cover the methods utilized, the data and tools used to achieve these goals. The proceeding chapter contains the results of the work, and finally, we discuss the implications, strengths and flaws of the work, future work and end with conclusions.
